---
title: "Compare ANI across sites"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(fig.width = 12, fig.height = 7)
```

```{r, message = FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
library(readr)
library(tidyr)
library(tibble)
```

### Set database variable

We will consider sylph outputs from two different databases, the commonly used GTDB r226 database, and the larger GlobDB database. We will set an variable here to determine which database we want to analyze the output from. 

```{r}
# db <- "GTDB"
db <- "Glob_DB"
```

### Load and clean data 

Load .sylphmpa files

```{r load sylphmpa data, message = FALSE}
sylphmpa_names <- list.files(paste0("../data/", db, "_per_sample"))
sylphmpa_files <- paste0("../data/", db, "_per_sample/", sylphmpa_names)
n_files <- length(sylphmpa_files)
sylphmpa_list <- vector(mode = "list", length = n_files)
for (i in 1:n_files) {
  sylphmpa_list[[i]] <- read_tsv(sylphmpa_files[i], skip = 1, show_col_types = F)
}
names(sylphmpa_list) <- str_remove(sylphmpa_names, ".sylphmpa")
```

```{r make strain df}
strain_df <- bind_rows(sylphmpa_list, .id = "site") %>% 
  rename(ani = `ANI (if strain-level)`, # simplify names 
         coverage = `Coverage (if strain-level)`) %>%
  filter(!is.na(ani)) %>% # only keep strain level clades
  separate_wider_delim(site, names = c("sequencing", "site", "replicate", "long_read"), 
                       delim = "_", too_few = "align_start") %>% # separate site, replicate, and whether long read sequencing is performed 
  mutate(location = str_remove_all(site, "[0-9]")) # get location from site 
```

### Explore data 

Summary of strains found in samples. 

```{r strain_summary}
nrow(strain_df) # observations (paired sample and strain)
length(unique(strain_df$clade_name)) # unique strains found in dataset 
strain_summary <- strain_df %>%
  group_by(clade_name) %>%
  summarise(n_samples = n(),
            n_sites = n_distinct(site),
            n_locations = n_distinct(location)) 
summary(strain_summary$n_samples) # number of samples each strain appears in 
summary(strain_summary$n_sites) # number of sites each strain appears in 
summary(strain_summary$n_locations) # number of locations each strain appears in 
```

```{r, include = FALSE}
n_clade <- length(unique(strain_df$clade_name))
n_single_sample <- sum(strain_summary$n_samples == 1)
n_single_site <- sum(strain_summary$n_sites == 1)
n_single_location <- sum(strain_summary$n_locations == 1)
```

Out of the `r n_clade` strains, `r n_single_sample` (`r round(n_single_sample / n_clade, 2) * 100`%) appear in a single sample, `r n_single_site` (`r round(n_single_site / n_clade, 2) * 100`%) appear in a single site, and `r n_single_location` (`r round(n_single_location / n_clade, 2) * 100`%) appear in a single location (representing a state or city). We will limit our analysis to the `r n_clade - n_single_site` strains that appear in multiple sites. 

```{r}
multiple_site_strains <- strain_summary %>%
  filter(n_sites > 1) %>%
  pull(clade_name)
multiple_site_df <- strain_df %>%
  filter(clade_name %in% multiple_site_strains)
```

Look at distribution of mean ANI for each strain. 

```{r}
multiple_site_summary <- strain_df %>%
  group_by(clade_name) %>%
  summarise(n_sites = n_distinct(site),
            mean_ani = mean(ani, na.rm = TRUE),
            var_ani = var(ani, na.rm = TRUE))
multiple_site_summary %>%
  ggplot(aes(x = mean_ani)) + 
  geom_histogram() + 
  labs(y = "Count", x = "Mean ANI") +
  theme_minimal()
```

Look at distribution of ANI across sites.

```{r}
multiple_site_df %>%
  ggplot(aes(y = ani, x = site, group = site)) + 
  geom_boxplot() +
  labs(x = "Site", y = "ANI") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
multiple_site_df %>%
  ggplot(aes(y = ani, x = location, group = location)) + 
  geom_boxplot() +
  labs(x = "Location", y = "ANI") + 
  theme_minimal() 
  
 # Order by median - might be helpful for identifying outliers
multiple_site_df %>%
  ggplot(aes(y = ani, x = forcats::fct_reorder(site, ani, median), group = site)) + 
  geom_boxplot() +
  labs(x = "Site", y = "ANI") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
multiple_site_df %>%
  ggplot(aes(y = ani, x = forcats::fct_reorder(location, ani, median), group = location)) + 
  geom_boxplot() +
  labs(x = "Location", y = "ANI") + 
  theme_minimal() 
```

Identify 20 strains that are present in the most number of sites, make heatmap with ANI across sites.

```{r, fig.height = 10}
n <- 20
# determine n clades with the highest prevalence 
top_n_clades <- strain_summary %>%
  mutate(clade_name = str_remove(clade_name, "^.*?(?=s__)") %>%
           str_replace("(?=t__)", "\n") %>%
           str_replace_all(" ", "\n")) %>%
  slice_max(order_by = n_sites, n = n) %>%
  pull(clade_name)
# alphabetical ordering of sites (for plotting)
site_levels <- sort(unique(multiple_site_df$site))
# df for top n strains across all sites 
top_n_df <- multiple_site_df %>%
  group_by(clade_name, site) %>%
  summarise(mean_site_ani = mean(ani),
            .groups = "drop_last") %>% # mean ani, over site and strain 
  mutate(clade_name = str_remove(clade_name, "^.*?(?=s__)") %>%
           str_replace("(?=t__)", "\n") %>%
           str_replace_all(" ", "\n")) %>% # only show species and strain 
  filter(clade_name %in% top_n_clades) %>%
  mutate(clade_name = factor(clade_name, levels = top_n_clades),
         site = factor(site, levels = site_levels))

# df for mean ANI by site (across all strains in multiple sites)
site_summ_df <- multiple_site_df %>%
  group_by(site) %>%
  summarise(mean_site_ani = mean(ani)) %>%
  mutate(site = factor(site, levels = site_levels)) %>%
  filter(site %in% unique(top_n_df$site)) %>%
  mutate(clade_name = "all strains in\nmultiple sites")
  
# df for mean ANI by strain (across all sites)
strain_summ_df <- multiple_site_df %>%
  group_by(clade_name) %>%
  summarise(mean_site_ani = mean(ani)) %>%
  mutate(clade_name = str_remove(clade_name, "^.*?(?=s__)") %>%
           str_replace("(?=t__)", "\n") %>%
           str_replace_all(" ", "\n")) %>%
  filter(clade_name %in% top_n_clades) %>%
  mutate(clade_name = factor(clade_name, levels = top_n_clades)) %>%
  mutate(site = "all sites")

# combine dfs 
full_df <- rbind(top_n_df, site_summ_df, strain_summ_df) %>%
  mutate(site = factor(site, levels = c(site_levels, "all sites")), 
         clade_name = factor(clade_name, levels = c(top_n_clades, "all strains in\nmultiple sites")))

# plot heatmap 
heatmap <- full_df %>%
  ggplot(aes(x = site, y = clade_name, fill = mean_site_ani)) + 
  geom_tile() +
  scale_fill_viridis_c(limits = c(95, 100)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "bottom") +
  labs(x = "Site", y = "Strain (twenty most prevalent)", fill = "Mean site ANI")

heatmap
``` 

Hierarchical clustering of sites based on presence/absence of strains.

```{r}
# using just the top n strains
ani_mat <- top_n_df %>%
  pivot_wider(names_from = clade_name, values_from = mean_site_ani) %>%
  column_to_rownames("site") %>%
  mutate(across(everything(), ~ replace_na(.x, 0)))
pres_mat <- (ani_mat > 0) * 1
pres_dist <- dist(pres_mat)
clusters <- hclust(pres_dist)
plot(clusters)

# using all strains
ani_mat <- strain_df %>%
  group_by(site, clade_name) %>%
  summarise(mean_site_ani = mean(ani)) %>% 
  pivot_wider(names_from = clade_name, values_from = mean_site_ani) %>%
  column_to_rownames("site") %>%
  mutate(across(everything(), ~ replace_na(.x, 0)))
pres_mat <- (ani_mat > 0) * 1
pres_dist <- dist(pres_mat)
clusters <- hclust(pres_dist)
plot(clusters)
```

In both of these plots, we can see sites from locations E, TC, F, ND, and SC cluster, which matches the patterns that we see in the heatmap.

Now look into individual strains that appear in multiple sites, and compare ANI across sites. For this, I'm not sure what strains we want to look into, but if there are particular strains of interest then we could compare ANI values across sites. As an example, I'll start with the strain with the highest prevalence in the data.

```{r study strain}
# example: look at most prevalent strain 
strain <- strain_summary %>%
  slice_max(order_by = n_sites, n = 1) %>%
  pull(clade_name)

strain_df %>%
  filter(clade_name == strain) %>% # filter to just this strain 
  ggplot(aes(y = ani, x = site)) + 
  geom_point() + 
  labs(x = "Site", y = "ANI") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

strain_df %>%
  filter(clade_name == strain) %>%
  ggplot(aes(y = ani, x = forcats::fct_reorder(location, ani, median))) + 
  geom_boxplot() +
  labs(x = "Location", y = "ANI") +
  theme_minimal()
```
