---
title: "Sylph output analysis"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r load packages, include = FALSE}
library(dplyr)
library(ggplot2)
library(ggpubr)
library(stringr)
```

### Load and clean data 

Load the sylph output data. This code currently reads the short read output, but replacing line 19 with line 20 will read the long read output instead. 

```{r load data}
# Read in filename metadata 
file_metadata <- read.csv("../data/sylph_file_metadata.csv") # short read output
# file_metadata <- read.csv("../data/sylph_file_metadata_long_read.csv") # use this for long read output

file_list <- file_metadata$file_name

# Initialize a list to store the data frames, using file names 
data_frames <- vector(mode = "list", length = length(file_list))
names(data_frames) <- file_list

# Loop through each file in the file list
for (file_name in file_list) {
  # Construct the full path to the file
  file_path <- file.path("../data/", file_name)
  
  # Get column names
  columns <- str_split(readLines(file_path)[2], pattern = '\t')[[1]]
  
  # Read the file into a data frame, specifying the separator and column names
  df <- read.table(file_path, sep = "\t", col.names = columns, header = FALSE, fill = TRUE)
  
  # Add the data frame to the list, using the file name (without extension) as the key
  data_frames[[file_name]] <- df
}

# Print the number of files processed
cat("Number of files processed:", length(data_frames), "\n")
```

There are two different sets of files, paired files that start with "HHGCVDRX3" and use "L001" and "L002" to denote the same samples processed through different lanes, and single files that start with "tax_profile_merged". These files have slightly different clade names, so we will fix that. 

```{r clean variable name for clade name}
head(file_list) # file names for paired files
tail(file_list) # file names for pre-merged files
paired_files <- which(str_detect(file_list, "HHGCVDRX3"))
merged_files <- which(str_detect(file_list, "tax_profile"))

names(data_frames[[paired_files[1]]]) # variable names for data from paired files
names(data_frames[[merged_files[1]]]) # variable names for data from merged files

data_frames[paired_files] <- lapply(data_frames[paired_files],
                                    function(x) {
                                      x %>% rename(clade_name = X.clade_name)
                                    })
names(data_frames[[paired_files[1]]]) # now clade names for all elements of data_frames are the same
```

The sylph output includes a row for each taxonomic level found in the sample. We want to perform an analysis at the order level, so we have to remove all rows that correspond to broader or finer taxonomic levels. 

```{r trim data}
# for each element in data_frames, make a new element in trimmed_data that only retains rows at the order level
trimmed_data <- lapply(data_frames, function(x) {
  x %>% 
    filter(str_detect(clade_name, "o_"), # keep rows that have order, denoted by "o__"
           !str_detect(clade_name, "f_")) # remove rows that have family, denoted by "f__"
})
```

Now that we have data at the order level, we will remove other taxonomic information and remove the "o__" prefix. 

```{r clean taxonomy}
trimmed_data <- lapply(trimmed_data, function(x) {
  x %>%
    mutate(clade_name = str_replace(clade_name,".*?o__", "" )) # remove taxonomy info up to and including "o__" 
})
```

Next, we will combine list objects from pairs of objects with the same file names. Some samples were run across multiple lanes (e.g., have both "L001" to denote lane 1 and "L002" to denote lane 2), and we want to combine these. Other samples were combined prior to sylph analysis. 

```{r combine site data}
# Create a base identifier for each site by removing 'L001' and 'L002' from the output name
base_identifiers <- unique(str_remove(names(trimmed_data), "L00[12]"))
# Initialize an empty list with length of the number o f base identifiers
combined_data <- vector(mode = "list", length = length(base_identifiers))
names(combined_data) <- base_identifiers

# Loop through each key in the trimmed_data list
for (key in names(trimmed_data)) {
  # Get the base identifier for this output
  base_identifier <- str_remove(key, "L00[12]")
  
  # Combine the current data frame with the existing one for the base identifier
  combined_data[[base_identifier]] <- rbind(combined_data[[base_identifier]], trimmed_data[[key]])
}
```

For each sample, we will group by order and sum relative abundances and sequence abundances. 

```{r average site data}
combined_data <- lapply(combined_data, function(x) {
  x %>% 
    group_by(clade_name) %>% # group by order 
     mutate( # force abundances to be numeric variables 
      relative_abundance = as.numeric(relative_abundance),
      sequence_abundance = as.numeric(sequence_abundance)
    ) %>%
    summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>% # sum abundances at order level 
    ungroup()
})

paired_files <- which(str_detect(names(combined_data), "HHGCVDRX3"))
# for paired files, divide abundance sums by 2 
combined_data[paired_files] <- lapply(combined_data[paired_files], function(x) {
  x %>%
    mutate(across(contains("abundance"), ~ .x / 2))
})
```

We will combine the sylph output with environmental variables (soil and site properties). 

```{r add soil and site data}
# convert list to dataframe, converting list element names to id column
sylph_data <- bind_rows(combined_data, .id = "id")

# read in metal data 
metal_data <- readr::read_csv("../data/BioDIGS_20250220.csv")

# clean metal data 
metal_data <- 
  metal_data %>% 
  # select metal related variables 
  select(site_id, replicate, ends_with("EPA3051"), ends_with("Mehlich3"), "water_pH", "OM_by_LOI_pct") %>% 
  filter(As_EPA3051 != "Not applicable" & As_EPA3051 != "Not yet tested") %>% 
  # As can't be detected lower than 3.0
  mutate(As_EPA3051 = case_when(As_EPA3051 == "< 3.0" ~ "0", As_EPA3051 == "<3.0" ~ "0", TRUE ~ As_EPA3051)) %>%
  # Cd can't be detected lower than 0.2
  mutate(Cd_EPA3051 = case_when(Cd_EPA3051 == "< 0.2" ~ "0", Cd_EPA3051 == "<0.2" ~ "0", TRUE ~ Cd_EPA3051)) %>% 
  # Cu edge case
  mutate(Cu_EPA3051 = case_when(Cu_EPA3051 == "< 2.0" ~ "0", TRUE ~ Cu_EPA3051)) %>% 
  # aggregate to site level 
  group_by(site_id) %>% 
  # average metal data at the site level 
  summarize(across(
    !c(replicate),
    ~ mean(as.numeric(.x))
  ))

# read in climate data 
clim_data <- read.csv(file = "../data/site_soil_features_t.csv")
# only retain a subset of climate variables 
clim_vars_to_keep <- c("X", "mgmt_type", "vap", "pre", "cld", "tmp", "imperv")
clim_data <- clim_data %>% dplyr::select(tidyselect::all_of(clim_vars_to_keep)) %>% 
  rename(site_id = X)

# combine sylph data with metal and climate data 
full_data <- sylph_data %>% 
  mutate(id = stringr::str_replace(id,"__", "_L001_")) %>% # add lane back into paired ids
  left_join(file_metadata, by = c("id" = "file_name")) %>% # join with file metadata to include site_id
  left_join(metal_data, by = "site_id") %>% # join with metal data by site_id
  left_join(clim_data, by = "site_id") # join with climate data by site_id
```

### Plot data 

First, generate a list of environmental variables to investigate. 

```{r make variable list}
# relevant environmental data 
var_list <- c(
  colnames(full_data)[stringr::str_detect(colnames(full_data), "EPA3051")], # Heavy metal testing protocol starts with this
  colnames(full_data)[stringr::str_detect(colnames(full_data), "Mehlich3")], # Soil elements testing protocol starts with this
  "water_pH",
  "mgmt_type",
  "vap",
  "pre",
  "cld",
  "tmp",
  "imperv" 
)
```

Next, aggregate samples to the level of site id, and calculate relative abundances of orders for each site id. 

```{r community data}
community_data <-
  full_data %>% 
  group_by(site_id, clade_name) %>% # aggregate at the order and sampling site level
    summarize(across(
      c(relative_abundance, all_of(var_list)),
      ~ mean(.x) # calculate average abundances and environmental variables for each site
    )) %>%
    group_by(site_id) %>%
  mutate(mult_factor = sum(relative_abundance)) %>% 
  ungroup() %>% 
  mutate(relative_abundance = relative_abundance / mult_factor) %>% # calculate aggregated relative abundances 
  select(site_id, clade_name, relative_abundance) # only retain site, clade, and relative abundance 

# make a wide version of this data frame, rows are sites and columns are clade abundances 
community_data_wide <- 
  community_data %>% 
  tidyr::pivot_wider(names_from = clade_name, values_from = relative_abundance) %>% 
  arrange(site_id) %>% 
  tibble::column_to_rownames("site_id")
# replace NA with zero for orders that were not detected at each site 
community_data_wide[is.na(community_data_wide)] <- 0
```
  
Create an environmental data frame, with one row per site.  
  
```{r environmental data}
# Create environmental data frame 
clim_data_perm <- 
  full_data %>% 
  tidyr::drop_na(site_id) %>%  # remove rows missing site id
  dplyr::filter(site_id != "") %>% 
  select(!c(clade_name, relative_abundance, sequence_abundance, ANI..if.strain.level., id, date, root, seq_id, site_rep_id)) %>% # remove sequencing related variables 
  distinct() %>% # keep only unique rows, one per site 
  arrange(site_id) %>% 
  tibble::column_to_rownames("site_id") 
```

Create lists of three types of environmental variables. 

```{r environmental variable subsets}
heavy_metal_list <- colnames(full_data)[stringr::str_detect(colnames(full_data), "EPA3051")] # heavy metal testing
elemental_list <- colnames(full_data)[stringr::str_detect(colnames(full_data), "Mehlich3")] # soil element testing
environmental <- c( # other environmental variables
  "OM_by_LOI_pct",
  "water_pH",
  "mgmt_type",
  "vap",
  "pre",
  "cld",
  "tmp",
  "imperv" 
)
```

Calculate Bray-Curtis distances between samples based on sets of variables and do principal components analysis of distances. 

```{r pcoas}
# pcoa on distances from order abundances 
bray_curtis_dist_comm <- vegan::vegdist(community_data_wide, method = "bray")
bray_curtis_pcoa_comm <- ecodist::pco(bray_curtis_dist_comm)

# pcoa on distances from heavy metal measurements
bray_curtis_dist_heavy <- vegan::vegdist(clim_data_perm %>% select(all_of(heavy_metal_list)), method = "bray")
bray_curtis_pcoa_heavy <- ecodist::pco(bray_curtis_dist_heavy)

# pcoa on distances from soil element measurements 
bray_curtis_dist_elem <- vegan::vegdist(clim_data_perm %>% select(all_of(elemental_list)), method = "bray")
bray_curtis_pcoa_elem <- ecodist::pco(bray_curtis_dist_elem)

# pcoa on distances from other environmental measurements
bray_curtis_dist_envt <- vegan::vegdist(clim_data_perm %>% select(all_of(environmental)), method = "bray")
bray_curtis_pcoa_envt <- ecodist::pco(bray_curtis_dist_envt)

# data frame of first two to three vectors from each pcoa 
bray_curtis_pcoa_df <- data.frame(pcoa1 = bray_curtis_pcoa_comm$vectors[, 1], 
                                  pcoa2 = bray_curtis_pcoa_comm$vectors[, 2],
                                  pcoa3 = bray_curtis_pcoa_comm$vectors[, 3],
                                  heavy1 = bray_curtis_pcoa_heavy$vectors[, 1],
                                  heavy2 = bray_curtis_pcoa_heavy$vectors[, 2],
                                  elem1 = bray_curtis_pcoa_elem$vectors[, 1],
                                  elem2 = bray_curtis_pcoa_elem$vectors[, 2],
                                  envt1 = bray_curtis_pcoa_envt$vectors[, 1],
                                  envt2 = bray_curtis_pcoa_envt$vectors[, 2],
                                  clim_data_perm
)

# plot order abundance pcoa, colored by pH 
ggplot(data = bray_curtis_pcoa_df, aes(x=pcoa1, y=pcoa2, color = water_pH)) +
  geom_point(size = 5) +
  labs(x = "PC1",
       y = "PC2",
       color = "Water pH") +
  scale_color_viridis_c(end = .7) +
  theme(title = element_text(size = 10)) # makes titles smaller
ggsave("ph.png", height = 5, width = 6)

# plot order abundance pcoa, colored by temperature 
ggplot(data = bray_curtis_pcoa_df, aes(x=pcoa1, y=pcoa2, color = tmp)) +
  geom_point(size = 5) +
  labs(x = "PC1",
       y = "PC2",
       color = "Temperature") +
  scale_color_viridis_c(end = .7) +
  theme(title = element_text(size = 10)) 
ggsave("tmp.png", height = 5, width = 6)

# plot order abundance pcoa, colored by first coordinate of heavy metals pcoa
ggplot(data = bray_curtis_pcoa_df, aes(x=pcoa1, y=pcoa2, color = heavy1)) +
  geom_point(size = 5) +
  labs(x = "PC1",
       y = "PC2",
       color = "Heavy\nmetals\nPC 1") +
  scale_color_viridis_c(end = .7) +
  theme(title = element_text(size = 10)) 
ggsave("heavy1.png", height = 5, width = 6)
```

Perform permanovas to test whether variation in Bray–Curtis dissimilarities of order-level community composition is associated with environmental variables.

```{r permanovas}
# model with pH, temperature, and the first principal coordinate from heavy metals pcoa
permanova_ph <- vegan::adonis2(community_data_wide ~ water_pH + tmp + heavy1, data = bray_curtis_pcoa_df, permutations = 9999)
# model with temperature 
permanova_tmp <- vegan::adonis2(community_data_wide ~ tmp, data = bray_curtis_pcoa_df, permutations = 9999)
# model with first principal coordinate from heavy metals pcoa
permanova_heavy1 <- vegan::adonis2(community_data_wide ~ heavy1, data = bray_curtis_pcoa_df, permutations = 9999)
```

Generate a heatmap of bacterial order abundances. 

```{r heatmap of bacterial orders, fig.width = 15, fig.height = 9}
# plot relative abundances of all detected orders in each site 
ggplot(community_data, aes(
  forcats::fct_reorder(clade_name, relative_abundance, .desc = T, .fun = sum), 
  site_id, fill= relative_abundance)) + 
  geom_tile() +
  labs(y = "Sampling site", x = "Bacterial order", fill = "Relative\nabundance") +
  scale_fill_viridis_c() +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) 
ggsave("figures/heatmap_ordered.png", width = 15, height = 9)
```

Generate a heatmap of environmental variables. 

```{r heatmap of environmental variables, fig.width = 15, fig.height = 9}
clim_data_perm_plot <- 
  clim_data_perm %>%
  scale() %>% # standardize variables to have mean 0 and standard deviation of 1 
  as.data.frame() %>% 
  tibble::rownames_to_column("site_id") %>% 
  tidyr::pivot_longer(!c(site_id), names_to = "name", values_to = "value") # create long data frame with environmental variable name and value pairs 

clim_data_perm_plot$name <- factor( # make environmental variable a factor and reorder levels
  x = clim_data_perm_plot$name,
  levels = c("As_EPA3051", "Cd_EPA3051", "Cr_EPA3051", "Cu_EPA3051", "Ni_EPA3051", "Pb_EPA3051", "Zn_EPA3051",
             "P_Mehlich3","K_Mehlich3","Ca_Mehlich3","Mg_Mehlich3", "Mn_Mehlich3", "Zn_Mehlich3","Cu_Mehlich3","Fe_Mehlich3","B_Mehlich3","S_Mehlich3","Na_Mehlich3","Al_Mehlich3",
             "water_pH", "OM_by_LOI_pct","mgmt_type", "vap","pre" ,"cld","tmp","imperv" ))


# plot standardized environmental variable values by site
ggplot(clim_data_perm_plot, aes(
  name,
  site_id, fill= value)) + 
  geom_tile() +
  labs(y = "Sampling site", x = "Environmental variable", fill = "Value") + 
  scale_fill_viridis_c() +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) 
ggsave("figures/heatmap_envt_ordered.png", width = 15, height = 9)
```

Calculate observed Shannon diversity index and richness for each site, and plot Shannon diversity index by environmental variables.

```{r richness and heavy metals, fig.width = 15, fig.height = 9}
plot_dat <- 
  data.frame(div_shannon = vegan::diversity(community_data_wide, index = "shannon")) %>% # calculate Shannon diversity index
  tibble::rownames_to_column() %>% 
  left_join(tibble::rownames_to_column(clim_data_perm)) # add in environmental variables 

plot_dat <-
  plot_dat %>%
  left_join(data.frame(richness = vegan::specnumber(community_data_wide)) %>% # calculate richness
              tibble::rownames_to_column())

plot_dat <- plot_dat %>% tidyr::pivot_longer(all_of(var_list)) # make long data frame

ggplot(data = plot_dat, aes(x = log(value), y = div_shannon)) + # y is shannon diversity, x is log of environmental variable
  geom_point() +
  geom_smooth(method = "glm") + # add trend line 
  facet_wrap(~name, scales = "free_x") + # separate plots by environmental variable 
  labs(x = "Environmental measurement (log scale)", y = "H' (Shannon Diversity)") +
  theme_minimal(base_size = 13)

```

Next, we generate stacked bar charts of order relative abundances across sites. 

First, keep all orders and plot at the site replicate level.

```{r stacked bar chart}
ggplot(
  full_data,
  aes(
    x = site_rep_id,
    y = relative_abundance * 2,
    fill = clade_name
  )
) +
  geom_bar(stat = "identity", position = 'stack') +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  scale_fill_discrete(guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90,
    hjust = 1,
    vjust = .5,
    size = 7
  )) 
```

Next, aggregate by site. 

```{r bar chart aggregated by site}
ggplot(
  community_data,
  aes(
    x = site_id,
    y = relative_abundance * 2,
    fill = clade_name
  )
) +
  geom_bar(stat = "identity", position = 'stack') +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  scale_fill_discrete(guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90,
    hjust = 1,
    vjust = .5,
    size = 7
  )) 
```

Finally, aggregate by site and only keep the ten orders with the highest average relative abundances across sites. Collapse all other orders into an "other" category. 

```{r bar chart top n orders}
n_orders <- 10
top_n_clades <- community_data %>%
  group_by(clade_name) %>% # aggregate over site 
  summarise(sum_rel_abundance = sum(relative_abundance, na.rm = TRUE), .groups = "drop") %>% # sum of relative abundances over all sites 
  arrange(desc(sum_rel_abundance)) %>%
  slice_head(n = n_orders) %>% # get ten orders with the largest average relative abundance
  pull(clade_name)
community_data_top_n <- community_data %>%
  mutate(clade_name = ifelse(clade_name %in% top_n_clades, clade_name, "other")) %>% # if order isn't one of the top ten, recode it as "other" 
  group_by(site_id, clade_name) %>% # aggregate "other" category
  summarise(relative_abundance = sum(relative_abundance), .groups = "drop") # sum relative abundance across "other" category

order_levels <- community_data_top_n %>%
  filter(clade_name != "other") %>% # for top ten orders
  group_by(clade_name) %>%
  summarise(mean_rel_abundance = mean(relative_abundance, na.rm = TRUE), .groups = "drop") %>% # order by mean relative abundance
  arrange(mean_rel_abundance) %>%
  pull(clade_name)

community_data_top_n <- community_data_top_n %>%
  mutate( # make clade_name a factor to order by mean relative abundances, add "other" in first 
    clade_name = factor(clade_name, levels = c("other", order_levels))
  )

palette_rest <- scales::hue_pal()(length(order_levels))  # set color palette for plot
names(palette_rest) <- order_levels
fill_values <- c(other = "gray70", palette_rest) # set "other" to be gray

# make stacked bar plot
ggplot(
  community_data_top_n,
  aes(x = site_id, y = relative_abundance, fill = clade_name)
) +
  geom_col(position = "stack") +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  scale_fill_manual(values = fill_values, breaks = c(rev(order_levels), "other")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, size = 7)) 
```
