---
title: "Sylph output analysis"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r load packages, include = FALSE}
library(tidyverse)
library(ggpubr)
library(vegan)
library(ecodist)
```

### Set database variable

We will consider sylph outputs from two different databases, the commonly used GTDB r226 database, and the larger GlobDB database. We will set an variable here to determine which database we want to analyze the output from. 

```{r}
db <- "GTDB"
# db <- "GlobDB"
```

### Set soil and site data variables

```{r}
restricted <- FALSE # TRUE if using data for all sites
if (restricted) {
  fig_folder <- "../figures_restricted/" # if restricted, save figures one folder up, not git tracked
} else {
  fig_folder <- "figures/" # if not restricted, save as figures in this repo 
}
local <- FALSE # if FALSE pull site and soil data from GitHub, if TRUE pull locally from data folder one directory up 
snapshot <- "20260310" # snapshot of data we are pulling (if non-restricted)

if (restricted & (!local)) {
  stop("If using restricted data, it needs to be saved locally.")
}
```

### Load and clean data 

Load in the table of sylph output. Each row represents a taxon identified in the sylph run, and each column represents a sample. 

```{r load data}
sylph_output <- read_tsv(paste0("../data/merged_relative_abundance_", db, ".tsv"), col_names = TRUE)
names(sylph_output)[-1] <- str_remove(names(sylph_output)[-1], "^([^/]*/){8}") %>% # remove up through the first 8 "/" in file name, since these are uninformative or redundant
  str_remove(".fastq.gz") # remove file extension 
sylph_output <- sylph_output %>%
  pivot_longer(names(sylph_output)[-1], names_to = "sample_id", values_to = "rel_abund") %>% # wide to long format 
  separate_wider_delim(sample_id, names = c("platform", "sample_id"), delim = "/") %>% # split sequencing platform from sample id
  mutate(sample_id = str_remove(sample_id, "Molecular_")) %>% 
  separate_wider_delim(sample_id, names = c("site_id", "replicate", "read_type"), delim = "_", 
                       cols_remove = FALSE, too_few = "align_start") %>% # split sample id into site id, replicate, and read type
  select(-read_type) # remove read type (this is R1 for all short reads, NA for all long reads )
```

The sylph output includes a row for each taxonomic level found in the sample. We want to perform an analysis at the order level, so we have to remove all rows that correspond to broader or finer taxonomic levels. 

```{r trim data}
# only retain rows at the order level
trimmed_data <- sylph_output %>%
  filter(str_detect(clade_name, "o_"), # keep rows that have order, denoted by "o__"
           !str_detect(clade_name, "f_")) # remove rows that have family, denoted by "f__"
```

Now that we have data at the order level, we will remove other taxonomic information and remove the "o__" prefix. 

```{r clean taxonomy}
trimmed_data <- trimmed_data %>%
  mutate(clade_name = str_replace(clade_name,".*?o__", "" )) # remove taxonomy info up to and including "o__" 
```

For each sample, we will group by order and calculate mean relative abundances and sequence abundances. 
```{r average site data}
combined_data <- trimmed_data %>% 
  group_by(clade_name, platform, site_id, replicate, sample_id) %>% # aggregate over clade_name
  summarise(rel_abund_sum = sum(rel_abund, na.rm = TRUE), .groups = "drop") %>%
  mutate(read_type = ifelse(platform == "NovaSeq 6000", "short", "long"),
         site_read_id = paste(site_id, read_type, sep = "_"))
```

We will now read in soil and site data. 

```{r load soil and site data}
# read in site data and soil data
if (local) {
  if (restricted) {
    site_data <- read.csv("../data/data_restricted/BioDIGS Sample Data and Kit Request MASTER - ALL_SAMPLES - SITE DATA.csv")
    soil_data <- read.csv("../data/data_restricted/BioDIGS Sample Data and Kit Request MASTER - ALL_SAMPLES - SOIL SAMPLES.csv")
    seq_data <- read.csv("../data/data_restricted/BioDIGS Sample Data and Kit Request MASTER - ALL_SAMPLES - SEQ SAMPLES.csv")
  } else {
    site_data <- read.csv(paste0("../data/BioDIGS_sites_", snapshot, ".csv"), row.names = 1)
    soil_data <- read.csv(paste0("../data/BioDIGS_soil_", snapshot, ".csv"), row.names = 1)
    seq_data <- read.csv(paste0("../data/BioDIGS_seq_", snapshot, ".csv"), row.names = 1)
  }
} else {
  site_data <- read.csv(paste0("https://raw.githubusercontent.com/FredHutch/GDSCNsoilsites/refs/heads/main/data/snapshots/BioDIGS_sites_", snapshot, ".csv"), row.names = 1)
  soil_data <- read.csv(paste0("https://raw.githubusercontent.com/FredHutch/GDSCNsoilsites/refs/heads/main/data/snapshots/BioDIGS_soil_", snapshot, ".csv"), row.names = 1)
  seq_data <- read.csv(paste0("https://raw.githubusercontent.com/FredHutch/GDSCNsoilsites/refs/heads/main/data/snapshots/BioDIGS_seq_", snapshot, ".csv"), row.names = 1)
}
```

Clean the soil data and average measurements across replicates. 

```{r clean soil data}
soil_data <- soil_data %>%
  # select relevant soil variables
  select(site_id, replicate, ends_with("EPA3051"), ends_with("Mehlich3"), 
         "water_pH", "OM_by_LOI_pct", "vap", "pre", "cld", "tmp") %>%
  filter(As_EPA3051 != "Not applicable" & As_EPA3051 != "Not yet tested") %>%
  # As can't be detected lower than 3.0
  mutate(As_EPA3051 = case_when(As_EPA3051 == "< 3.0" ~ "0", As_EPA3051 == "<3.0" ~ "0", TRUE ~ As_EPA3051)) %>%
  # Cd can't be detected lower than 0.2
  mutate(Cd_EPA3051 = case_when(Cd_EPA3051 == "< 0.2" ~ "0", Cd_EPA3051 == "<0.2" ~ "0", Cd_EPA3051 == "< 0.20" ~ "0", TRUE ~ Cd_EPA3051)) %>%
  # Cu edge case
  mutate(Cu_EPA3051 = case_when(Cu_EPA3051 == "< 2.0" ~ "0", TRUE ~ Cu_EPA3051)) %>%
  # Pb edge case
  mutate(Pb_EPA3051 = as.character(Pb_EPA3051)) %>% 
  mutate(Pb_EPA3051 = case_when(Pb_EPA3051 == "< 2.0" ~ "0", TRUE ~ Pb_EPA3051)) %>%
  # when OM_by_LOI_pct is coded as ISS, set to NA
  mutate(OM_by_LOI_pct = case_when(OM_by_LOI_pct == "ISS" ~ NA, TRUE ~ OM_by_LOI_pct)) %>% 
  # for environmental variables with "sampling date not yet available", set as NA 
  mutate(across(c(vap, pre, cld, tmp), 
                ~ na_if(.x, "sampling date not yet available"))) %>% 
  # aggregate to site level
  group_by(site_id) %>%
  # average metal data at the site level
  summarize(across(
    !c(replicate),
    ~ mean(as.numeric(.x))
  ))
```

Clean site data.

```{r clean site data}
site_data <- site_data %>%
  mutate(
    # Recode rare management types 
    mgmt_type = case_when(
      mgmt_type %in% c(
        "Botanical Garden",
        "Mowed to control kudzu, but other than that unmanaged area"
      ) ~ "Managed: mowed/tended lawns, yard, or park",
      mgmt_type == "not yet provided" ~ NA,
      TRUE ~ mgmt_type
    )) %>%
  mutate(mgmt_type = as.numeric(as.factor(mgmt_type)))
```

Check on sequencing date in sequencing metadata.

```{r investigate sequencing date}
seq_data <- 
  seq_data %>%
  add_count(seq_date, name = "n_per_date") %>% # count number of samples sequenced per date
  mutate(seq_date_cat = ifelse(n_per_date < 5, "other", seq_date)) %>% # if less than 5 samples on that date, set category to "other"
  select(-n_per_date) %>%
  mutate(sequencing_instrument = ifelse(sequencing_instrument == FALSE, NA, sequencing_instrument),
         read_type = ifelse(sequencing_instrument == "NovaSeq 6000", "short", "long"))

# check on if any sites were sequenced on different days
seq_data %>%
  group_by(site_id, sequencing_instrument) %>%
  summarise(n_seq_dates = n_distinct(seq_date_cat)) %>%
  filter(n_seq_dates > 1)

# aggregate sequencing data to the site level 
seq_data <- seq_data %>%
  filter(!is.na(sequencing_instrument)) %>% 
  group_by(site_id, read_type) %>%
  summarise(seq_date_summ = ifelse(
    n_distinct(seq_date_cat, na.rm = TRUE) == 1,
    first(seq_date_cat), "multiple"
  ))
```

We now combine the sylph output with the sites and soil data. 

```{r join data}
# combine sylph data with metal and climate data 
full_data <- combined_data %>% 
  left_join(soil_data, by = "site_id") %>% # join with soil data that has been aggregated to the site level
  left_join(site_data, by = "site_id") %>% # join with site data 
  left_join(seq_data, by = c("site_id", "read_type")) # join with sequencing data
```

### Plot data 

First, aggregate samples to the level of site id, and calculate relative abundances of orders for each site id. 

```{r community data}
community_data <- full_data %>%
  group_by(site_id, read_type, site_read_id, clade_name) %>% # aggregate at order and sampling site level (and separate short reads and long reads)
  summarize(relative_abundance = mean(rel_abund_sum), # get average relative abundance of orders at each site
            .groups = "drop_last") %>% 
  mutate(mult_factor = sum(relative_abundance)) %>% 
  ungroup() %>%
  mutate(relative_abundance = relative_abundance / mult_factor) %>% # rescale relative abundances after aggregation 
  select(site_id, read_type, site_read_id, clade_name, relative_abundance)

# make a wide version of this data frame, rows are sites and columns are clade abundances 
community_data_wide <- 
  community_data %>% 
  select(-site_id, -read_type) %>% 
  pivot_wider(names_from = clade_name, values_from = relative_abundance) %>% 
  arrange(site_read_id) %>% 
  column_to_rownames("site_read_id")
# replace NA with zero for orders that were not detected at each site 
community_data_wide[is.na(community_data_wide)] <- 0
```
  
Create an environmental data frame, with one row per site.  
  
```{r environmental data}
# Create environmental data frame 
clim_data_perm <- 
  full_data %>% 
  tidyr::drop_na(site_id) %>%  # remove rows missing site id
  dplyr::filter(site_id != "") %>% 
  select(!c(clade_name, platform, read_type, site_read_id, replicate, sample_id, rel_abund_sum, seq_date_summ)) %>% # remove sequencing related variables 
  distinct() %>% # keep only unique rows, one per site 
  arrange(site_id) %>%
  column_to_rownames("site_id")
```

Create lists of three types of environmental variables. 

```{r environmental variable subsets}
heavy_metal_vars <- colnames(full_data)[str_detect(colnames(full_data), "EPA3051")] # heavy metal testing
elemental_vars <- colnames(full_data)[str_detect(colnames(full_data), "Mehlich3")] # soil element testing
environmental_vars <- c( # other environmental variables
  "OM_by_LOI_pct",
  "water_pH",
  "mgmt_type",
  "vap",
  "pre",
  "cld",
  "tmp",
  "pct_imp_2024" 
)
```

Calculate Bray-Curtis distances between samples based on sets of variables and do principal components analysis of distances. 

```{r pcoas}
# pcoa on distances from order abundances 
bray_curtis_dist_comm <- vegdist(community_data_wide, method = "bray")
bray_curtis_pcoa_comm <- pco(bray_curtis_dist_comm)

# ndms on distances from order abundances
set.seed(15)
nmds <- metaMDS(community_data_wide, distance = "bray", k = 2, trymax = 100)
nmds$stress # fair/good representation

# pcoa on distances from heavy metal measurements
bray_curtis_dist_heavy <- clim_data_perm %>%
  select(all_of(heavy_metal_vars)) %>%
  drop_na() %>% 
  vegdist(method = "bray")
bray_curtis_pcoa_heavy <- pco(bray_curtis_dist_heavy)

# pcoa on distances from soil element measurements 
bray_curtis_dist_elem <- clim_data_perm %>%
  select(all_of(elemental_vars)) %>%
  drop_na() %>% 
  vegdist(method = "bray")
bray_curtis_pcoa_elem <- pco(bray_curtis_dist_elem)

# pcoa on distances from other environmental measurements
bray_curtis_dist_envt <- clim_data_perm %>%
  select(all_of(environmental_vars)) %>%
  drop_na() %>% 
  vegdist(method = "bray")
bray_curtis_pcoa_envt <- pco(bray_curtis_dist_envt)

# data frame of first two to three vectors from each pcoa 
bray_curtis_pcoa_df <- data.frame(site_read_id = rownames(bray_curtis_pcoa_comm$vectors),
                                  pcoa1 = bray_curtis_pcoa_comm$vectors[, 1], 
                                  pcoa2 = bray_curtis_pcoa_comm$vectors[, 2],
                                  pcoa3 = bray_curtis_pcoa_comm$vectors[, 3]) %>%
  left_join(data.frame(scores(nmds)$sites) %>% rownames_to_column(var = "site_read_id")) %>% 
  mutate(site_id = str_remove(site_read_id, "_.*"),
         read_type = str_remove(site_read_id, "^.*_")) %>% 
  left_join(data.frame(site_id = rownames(bray_curtis_pcoa_heavy$vectors),
                       heavy1 = bray_curtis_pcoa_heavy$vectors[, 1],
                       heavy2 = bray_curtis_pcoa_heavy$vectors[, 2]), 
            by = "site_id") %>%
  left_join(data.frame(site_id = rownames(bray_curtis_pcoa_elem$vectors),
                       elem1 = bray_curtis_pcoa_elem$vectors[, 1],
                       elem2 = bray_curtis_pcoa_elem$vectors[, 2]),
            by = "site_id") %>%
  left_join(data.frame(site_id = rownames(bray_curtis_pcoa_envt$vectors),
                       envt1 = bray_curtis_pcoa_envt$vectors[, 1],
                       envt2 = bray_curtis_pcoa_envt$vectors[, 2]),
            by = "site_id") %>%
  left_join(clim_data_perm %>%
              rownames_to_column(var = "site_id"), by = "site_id") %>%
  left_join(full_data %>% select(site_read_id, seq_date_summ), by = "site_read_id")

# plot order abundance pcoa, colored by pH 
ggplot(data = bray_curtis_pcoa_df, aes(x=pcoa1, y=pcoa2, color = water_pH, shape = read_type)) +
  geom_point(size = 5) +
  labs(x = "PC1",
       y = "PC2",
       color = "Water pH", 
       shape = "Read type") +
  scale_color_viridis_c(end = .7) +
  scale_shape_manual(values = c(12, 16)) + 
  theme(title = element_text(size = 10)) # makes titles smaller
ggsave(paste0(fig_folder, "pcoa_colored_by_ph_", db, ".png"), height = 5, width = 6)

# plot order abundance pcoa, colored by temperature 
ggplot(data = bray_curtis_pcoa_df, aes(x=pcoa1, y=pcoa2, color = tmp, shape = read_type)) +
  geom_point(size = 5) +
  labs(x = "PC1",
       y = "PC2",
       color = "Temperature",
       shape = "Read type") +
  scale_color_viridis_c(end = .7) +
  scale_shape_manual(values = c(12, 16)) + 
  theme(title = element_text(size = 10)) 
ggsave(paste0(fig_folder, "pcoa_colored_by_temp_", db, ".png"), height = 5, width = 6)

# plot order abundance pcoa, colored by first coordinate of heavy metals pcoa
ggplot(data = bray_curtis_pcoa_df, aes(x=pcoa1, y=pcoa2, color = heavy1, shape = read_type)) +
  geom_point(size = 5) +
  labs(x = "PC1",
       y = "PC2",
       color = "Heavy\nmetals\nPC 1",
       shape = "Read type") +
  scale_color_viridis_c(end = .7) +
  scale_shape_manual(values = c(12, 16)) + 
  theme(title = element_text(size = 10)) 
ggsave(paste0(fig_folder, "pcoa_colored_by_heavy_metals_", db, ".png"), height = 5, width = 6)

# plot order abundance pcoa, colored by sequencing data
ggplot(data = bray_curtis_pcoa_df, aes(x=pcoa1, y=pcoa2, color = seq_date_summ, shape = read_type)) +
  geom_point(size = 5) +
  labs(x = "PC1",
       y = "PC2",
       color = "Sequencing date",
       shape = "Read type") +
  scale_shape_manual(values = c(12, 16)) + 
  theme(title = element_text(size = 10)) 
ggsave(paste0(fig_folder, "pcoa_colored_by_sequencing_date_", db, ".png"), height = 5, width = 6)

# plot order abundance nmds, colored by pH 
ggplot(data = bray_curtis_pcoa_df, aes(x=NMDS1, y=NMDS2, color = water_pH, shape = read_type)) +
  geom_point(size = 5) +
  labs(x = "NMDS1",
       y = "NMDS2",
       color = "Water pH", 
       shape = "Read type") +
  scale_color_viridis_c(end = .7) +
  scale_shape_manual(values = c(12, 16)) + 
  theme(title = element_text(size = 10)) # makes titles smaller
ggsave(paste0(fig_folder, "nmds_colored_by_ph_", db, ".png"), height = 5, width = 6)

# plot order abundance nmds, colored by temperature 
ggplot(data = bray_curtis_pcoa_df, aes(x=NMDS1, y=NMDS2, color = tmp, shape = read_type)) +
  geom_point(size = 5) +
  labs(x = "NMDS1",
       y = "NMDS2",
       color = "Temperature",
       shape = "Read type") +
  scale_color_viridis_c(end = .7) +
  scale_shape_manual(values = c(12, 16)) + 
  theme(title = element_text(size = 10)) 
ggsave(paste0(fig_folder, "nmds_colored_by_temp_", db, ".png"), height = 5, width = 6)

# plot order abundance nmds, colored by first coordinate of heavy metals pcoa
ggplot(data = bray_curtis_pcoa_df, aes(x=NMDS1, y=NMDS2, color = heavy1, shape = read_type)) +
  geom_point(size = 5) +
  labs(x = "NMDS1",
       y = "NMDS2",
       color = "Heavy\nmetals\nPC 1",
       shape = "Read type") +
  scale_color_viridis_c(end = .7) +
  scale_shape_manual(values = c(12, 16)) + 
  theme(title = element_text(size = 10)) 
ggsave(paste0(fig_folder, "nmds_colored_by_heavy_metals_", db, ".png"), height = 5, width = 6)

# plot order abundance nmds, colored by sequencing data
ggplot(data = bray_curtis_pcoa_df, aes(x=NMDS1, y=NMDS2, color = seq_date_summ, shape = read_type)) +
  geom_point(size = 5) +
  labs(x = "NMDS1",
       y = "NMDS2",
       color = "Sequencing date",
       shape = "Read type") +
  scale_shape_manual(values = c(12, 16)) + 
  theme(title = element_text(size = 10)) 
ggsave(paste0(fig_folder, "nmds_colored_by_sequencing_date_", db, ".png"), height = 5, width = 6)
```

Perform permanovas to test whether variation in Bray–Curtis dissimilarities of order-level community composition is associated with environmental variables.

```{r permanovas}
# model with pH, temperature, and the first principal coordinate from heavy metals pcoa
permanova_ph <- adonis2(community_data_wide ~ water_pH + tmp + heavy1, data = bray_curtis_pcoa_df, permutations = 9999, na.action = na.omit)
# model with temperature 
permanova_tmp <- adonis2(community_data_wide ~ tmp, data = bray_curtis_pcoa_df, permutations = 9999, na.action = na.omit)
# model with first principal coordinate from heavy metals pcoa
permanova_heavy1 <- adonis2(community_data_wide ~ heavy1, data = bray_curtis_pcoa_df, permutations = 9999, na.action = na.omit)
```

Generate a heatmap of order abundances. 

```{r heatmap of orders, fig.width = 15, fig.height = 9}
# plot relative abundances of all detected orders in each site 
community_data %>%
  filter(relative_abundance > 0) %>% 
  mutate(read_type = factor(read_type, levels = c("short", "long"),
                            labels = c("Short\nreads", "Long\nreads"))) %>%
  #arrange(read_type, site_id) %>%
  mutate(site_read_id = factor(site_read_id, levels = unique(site_read_id))) %>%
  ggplot(aes(
    fct_reorder(clade_name, relative_abundance, .desc = T, .fun = sum), 
    site_read_id, 
    fill= relative_abundance)) + 
  geom_tile() +
  labs(y = "Sampling site", x = "Order", fill = "Relative\nabundance") +
  scale_fill_viridis_c() +
  theme_minimal(base_size = 13) +
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 4),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 7)
  ) +
  facet_grid(rows = "read_type", scales = "free_y", space = "free") +
  scale_y_discrete(labels = function(x) sub("_.*", "", x))
ggsave(paste0(fig_folder, "heatmap_ordered_", db, ".png"), width = 15, height = 9)
```

Generate a heatmap of environmental variables. 

```{r heatmap of environmental variables, fig.width = 15, fig.height = 9}
vars <- c(heavy_metal_vars, elemental_vars, environmental_vars)
clim_data_perm_plot <- 
  clim_data_perm %>%
  select(all_of(vars)) %>% 
  scale() %>% # standardize variables to have mean 0 and standard deviation of 1 
  as.data.frame() %>% 
  rownames_to_column("site_id") %>% 
  pivot_longer(!c(site_id), names_to = "name", values_to = "value") # create long data frame with environmental variable name and value pairs 

clim_data_perm_plot$name <- factor( # make environmental variable a factor and reorder levels
  x = clim_data_perm_plot$name,
  levels = vars)

# plot standardized environmental variable values by site
ggplot(clim_data_perm_plot, aes(
  name,
  site_id, fill= value)) + 
  geom_tile() +
  labs(y = "Sampling site", x = "Environmental variable", fill = "Value") + 
  scale_fill_viridis_c() +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    axis.text.y = element_text(size = 7)
  ) 
ggsave(paste0(fig_folder, "heatmap_env_ordered.png"), width = 15, height = 9)
```

Calculate observed Shannon diversity index and richness for each site, and plot Shannon diversity index by environmental variables.

```{r richness and heavy metals, fig.width = 15, fig.height = 9, warning=FALSE}
plot_dat <- 
  data.frame(div_shannon = diversity(community_data_wide, index = "shannon")) %>% # calculate Shannon diversity index
  rownames_to_column(var = "site_read_id") %>% 
  mutate(site_id = str_remove(site_read_id, "_.*"),
         read_type = str_remove(site_read_id, "^.*_")) %>%
  left_join(rownames_to_column(clim_data_perm, var = "site_id"), 
            by = "site_id") # add in environmental variables 

plot_dat <-
  plot_dat %>%
  left_join(data.frame(richness = specnumber(community_data_wide)) %>% # calculate richness
              rownames_to_column(var = "site_read_id"), by = "site_read_id")

plot_dat <- plot_dat %>% pivot_longer(all_of(vars)) # make long data frame

ggplot(data = plot_dat, aes(x = log(value), y = div_shannon)) + # y is shannon diversity, x is log of environmental variable
  geom_point() +
  geom_smooth(method = "glm", formula = 'y ~ x') + # add trend line 
  facet_wrap(~name, scales = "free_x") + # separate plots by environmental variable 
  labs(x = "Environmental measurement (log scale)", y = "H' (Shannon Diversity)") +
  theme_minimal(base_size = 13)
ggsave(paste0(fig_folder, "shannon_env_", db, ".png"), width = 15, height = 9)
```

Next, we generate stacked bar charts of order relative abundances across sites. 

First, keep all orders and plot at the site replicate level.

```{r stacked bar chart}
full_data %>%
  group_by(sample_id, read_type, clade_name) %>% # aggregate at order and sampling site level
  summarize(relative_abundance = mean(rel_abund_sum), # get average relative abundance of orders at each site
            .groups = "drop_last") %>% 
  mutate(mult_factor = sum(relative_abundance)) %>% 
  ungroup() %>%
  mutate(relative_abundance = 100 * relative_abundance / mult_factor, # rescale relative abundances after aggregation 
         read_type = factor(read_type, levels = c("short", "long"),
                            labels = c("Short reads", "Long reads"))) %>% 
  ggplot(
    aes(
      x = sample_id,
      y = relative_abundance,
      fill = clade_name
    )
  ) +
  geom_bar(stat = "identity", position = 'stack') +
  labs(x = "Sample", y = "Relative Abundance", fill = "Order name") +
  scale_fill_discrete(guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90,
    hjust = 1,
    vjust = .5,
    size = 7
  )) +
  facet_grid(rows = NULL, cols = vars(read_type), scales = "free_x", space = "free")
ggsave(paste0(fig_folder, "rel_abund_by_sample_", db, ".png"), height = 9, width = 15)
```

Next, aggregate by site. 

```{r bar chart aggregated by site}
community_data %>% 
  mutate(read_type = factor(read_type, levels = c("short", "long"),
                            labels = c("Short reads", "Long reads"))) %>%
  ggplot(aes(
    x = site_read_id,
    y = relative_abundance * 100,
    fill = clade_name
  )
  ) +
  geom_bar(stat = "identity", position = 'stack') +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  scale_fill_discrete(guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90,
    hjust = 1,
    vjust = .5,
    size = 7
  )) +
  facet_grid(rows = NULL, cols = vars(read_type), scales = "free_x", space = "free") +
  scale_x_discrete(labels = function(x) sub("_.*", "", x))
ggsave(paste0(fig_folder, "rel_abund_by_site_", db, ".png"), height = 9, width = 15)
```

Make the same plot, ordering the sites by pH. 

```{r bar chart aggregated by site ordered by pH}
site_order <- community_data %>%
  left_join(soil_data %>% select(site_id, water_pH), by = "site_id") %>% 
  distinct(site_read_id, water_pH) %>%
  arrange(desc(water_pH)) %>%
  pull(site_read_id)
community_data %>% 
  mutate(read_type = factor(read_type, levels = c("short", "long"),
                            labels = c("Short reads", "Long reads")),
         site_read_id = factor(site_read_id, levels = site_order)) %>%
  ggplot(aes(
    x = site_read_id,
    y = relative_abundance * 100,
    fill = clade_name
  )
  ) +
  geom_bar(stat = "identity", position = 'stack') +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  scale_fill_discrete(guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90,
    hjust = 1,
    vjust = .5,
    size = 7
  )) +
  facet_grid(rows = NULL, cols = vars(read_type), scales = "free_x", space = "free") +
  scale_x_discrete(labels = function(x) sub("_.*", "", x))
#ggsave(paste0(fig_folder, "rel_abund_by_site_pH_", db, ".png"), height = 9, width = 15)
```

Make the same plot, ordering the sites by temperature. 

```{r bar chart aggregated by site ordered by temperature}
site_order <- community_data %>%
  left_join(soil_data %>% select(site_id, tmp), by = "site_id") %>% 
  distinct(site_read_id, tmp) %>%
  arrange(desc(tmp)) %>%
  pull(site_read_id)
community_data %>% 
  mutate(read_type = factor(read_type, levels = c("short", "long"),
                            labels = c("Short reads", "Long reads")),
         site_read_id = factor(site_read_id, levels = site_order)) %>%
  ggplot(aes(
    x = site_read_id,
    y = relative_abundance * 100,
    fill = clade_name
  )
  ) +
  geom_bar(stat = "identity", position = 'stack') +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  scale_fill_discrete(guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90,
    hjust = 1,
    vjust = .5,
    size = 7
  )) +
  facet_grid(rows = NULL, cols = vars(read_type), scales = "free_x", space = "free") +
  scale_x_discrete(labels = function(x) sub("_.*", "", x))
#ggsave(paste0(fig_folder, "rel_abund_by_site_tmp_", db, ".png"), height = 9, width = 15)
```

```{r bar chart aggregated by site comparing short and long reads}
community_data %>% 
  mutate(read_type = factor(read_type, levels = c("short", "long"),
                            labels = c("Short reads", "Long reads"))) %>%
  group_by(site_id) %>%
  filter(n_distinct(read_type) == 2) %>%
  ungroup() %>% 
  ggplot(aes(
    x = site_read_id,
    y = relative_abundance * 100,
    fill = clade_name
  )
  ) +
  geom_bar(stat = "identity", position = 'stack') +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  scale_fill_discrete(guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90,
    hjust = 1,
    vjust = .5,
    size = 7
  )) +
  scale_x_discrete(labels = function(x) sub("_", " ", x)) +
  facet_grid(cols = vars(site_id), scales = "free_x", space = "free")
ggsave(paste0(fig_folder, "rel_abund_by_site_long_v_short_", db, ".png"), height = 9, width = 15)
```

Finally, aggregate by site and only keep the ten orders with the highest average relative abundances across sites. Collapse all other orders into an "other" category. 

```{r bar chart top n orders}
n_orders <- 10
top_n_clades <- full_data %>%
  group_by(site_id, clade_name) %>% # aggregate at order and sampling site level
  summarize(relative_abundance = mean(rel_abund_sum), # get average relative abundance of orders at each site
            .groups = "drop_last") %>% 
  mutate(mult_factor = sum(relative_abundance)) %>% 
  ungroup() %>%
  mutate(relative_abundance = relative_abundance / mult_factor) %>%
  group_by(clade_name) %>% # aggregate over site 
  summarise(sum_rel_abundance = sum(relative_abundance, na.rm = TRUE), .groups = "drop") %>% # sum of relative abundances over all sites 
  arrange(desc(sum_rel_abundance)) %>%
  slice_head(n = n_orders) %>% # get n_orders orders with the highest average relative abundance
  pull(clade_name)
community_data_top_n <- community_data %>%
  mutate(clade_name = ifelse(clade_name %in% top_n_clades, clade_name, "other")) %>% # if order isn't one of the top ten, recode it as "other" 
  group_by(site_read_id, site_id, read_type, clade_name) %>% # aggregate "other" category
  summarise(relative_abundance = 100 * sum(relative_abundance), .groups = "drop") # sum relative abundance across "other" category

order_levels <- community_data_top_n %>%
  filter(clade_name != "other") %>% # for top ten orders
  group_by(clade_name) %>%
  summarise(mean_rel_abundance = mean(relative_abundance, na.rm = TRUE), .groups = "drop") %>% # order by mean relative abundance
  arrange(mean_rel_abundance) %>%
  pull(clade_name)

community_data_top_n <- community_data_top_n %>%
  mutate( # make clade_name a factor to order by mean relative abundances, add "other" in first 
    clade_name = factor(clade_name, levels = c("other", order_levels))
  ) %>%
  mutate(read_type = factor(read_type, levels = c("short", "long"),
                            labels = c("Short reads", "Long reads"))) 

palette_rest <- scales::hue_pal()(length(order_levels))  # set color palette for plot
names(palette_rest) <- order_levels
fill_values <- c(other = "gray70", palette_rest) # set "other" to be gray

# make stacked bar plot
ggplot(
  community_data_top_n,
  aes(x = site_read_id, y = relative_abundance, fill = clade_name)
) +
  geom_col(position = "stack") +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  scale_fill_manual(values = fill_values, breaks = c(rev(order_levels), "other")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, size = 7)) +
  facet_grid(rows = NULL, cols = vars(read_type), scales = "free_x", space = "free") +
  scale_x_discrete(labels = function(x) sub("_.*", "", x))
ggsave(paste0(fig_folder, "top_rel_abund_by_site_", db, ".png"), height = 9, width = 15)
```

Make the same plot, with sites ordered by pH.

```{r bar chart top n orders ordered by pH}
site_order <- community_data %>%
  left_join(soil_data %>% select(site_id, water_pH), by = "site_id") %>% 
  distinct(site_read_id, water_pH) %>%
  arrange(desc(water_pH)) %>%
  pull(site_read_id)

# make stacked bar plot
community_data_top_n %>%
  mutate(site_read_id = factor(site_read_id, levels = site_order)) %>%
  ggplot(aes(x = site_read_id, y = relative_abundance, fill = clade_name)) +
  geom_col(position = "stack") +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  scale_fill_manual(values = fill_values, breaks = c(rev(order_levels), "other")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, size = 7)) +
  facet_grid(rows = NULL, cols = vars(read_type), scales = "free_x", space = "free") +
  scale_x_discrete(labels = function(x) sub("_.*", "", x))
#ggsave(paste0(fig_folder, "top_rel_abund_by_site_pH_", db, ".png"), height = 9, width = 15)
```

Make the same plot, with sites ordered by temperature.

```{r bar chart top n orders ordered by temperature}
site_order <- community_data %>%
  left_join(soil_data %>% select(site_id, tmp), by = "site_id") %>% 
  distinct(site_read_id, tmp) %>%
  arrange(desc(tmp)) %>%
  pull(site_read_id)

# make stacked bar plot
community_data_top_n %>%
  mutate(site_read_id = factor(site_read_id, levels = site_order)) %>%
  ggplot(aes(x = site_read_id, y = relative_abundance, fill = clade_name)) +
  geom_col(position = "stack") +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  scale_fill_manual(values = fill_values, breaks = c(rev(order_levels), "other")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, size = 7)) +
  facet_grid(rows = NULL, cols = vars(read_type), scales = "free_x", space = "free") +
  scale_x_discrete(labels = function(x) sub("_.*", "", x))
#ggsave(paste0(fig_folder, "top_rel_abund_by_site_tmp_", db, ".png"), height = 9, width = 15)
```

```{r bar chart top orders aggregated by site comparing short and long reads}
community_data_top_n %>% 
  group_by(site_id) %>%
  filter(n_distinct(read_type) == 2) %>%
  ungroup() %>% 
  ggplot(aes(
    x = site_read_id,
    y = relative_abundance,
    fill = clade_name
  )
  ) +
  geom_bar(stat = "identity", position = 'stack') +
  labs(x = "Site Location", y = "Relative Abundance", fill = "Order name") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90,
    hjust = 1,
    vjust = .5,
    size = 7
  )) +
  scale_fill_manual(values = fill_values, breaks = c(rev(order_levels), "other")) +
  scale_x_discrete(labels = function(x) sub("_", " ", x)) +
  facet_grid(cols = vars(site_id), scales = "free_x", space = "free")
ggsave(paste0(fig_folder, "top_rel_abund_by_site_long_v_short_", db, ".png"), height = 9, width = 15)
```
